---
title: "Final Project:Milestones 6, Flu outbreak (simulated) and vaccine uptake in California"
author: "Ngu Claudia Ngeha, Ukah Chrisntus Eweh"
date: "`r Sys.Date()`"
output:
  html_document: default
  
---
```{r, include=FALSE}
library(readr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(data.table)
library(knitr)
library(stringr)
library(kableExtra)
library(DT)
library(png)
library(ggrepel)
```
##Rpubs link: https://rpubs.com/chrisukah/1127732

## Research Question

“Is there any correlation between COVID-19 vaccination rates on flu rates and/or severity ?”

## Introduction & Problem Statement

Seasonal influenza (the flu) is an acute respiratory infection caused by influenza viruses. There are around a billion cases of seasonal influenza annually, including 3–5 million cases of severe illness.It causes 290 000 to 650 000 respiratory deaths annually(WHO. 2023).It is common in all parts of the world. Most people recover without treatment. Hence it is of public health importance.Influenza spreads easily between people when they cough or sneeze. Vaccination is the best way to prevent the disease.Symptoms of influenza include acute onset of fever, cough, sore throat, body aches and fatigue.Influenza virus, commonly known as “Flu virus” is highly contagious via droplet transmission, and it belongs to the Orthomyxoviridae family. The two main influenza viruses that affect humans are A and B, where Influenza A viruses emerge seasonally from zoonotic reservoirs . There have been five known pandemics due to influenza, where the most recent one was in 2009 and 200,000 deaths were attributed to influenza. The population at higher risk of morbidity and mortality are the extreme ages of life, the pediatric population and the elderly during the flu season (that vary geographically) (Peteranderl et al., 2016). In the US, the estimated incidence of flu among seasons prior to COVID-19 pandemic was from 3% - 11% in all ages, and 9.3% for the pediatric population 0-17 years old, (okars et al., 2017). The statistical estimate of influenza incidence in the US among all ages ranged from 3.0%–11.3% among seasons (Tokars et al., 2018).Flu vaccination coverage among the pediatric population in the US during 2019-2022 coverage was around 62%  (CDC, 2023). On the other hand, COVID-19 vaccination coverage was strengthened in the pediatric population during 2022 reaching a national wide coverage of 50% approximately (Nguyen et al., 2023). While flu cases were increasing as well as COVID-19 vaccination coverage, we intend to determine if there is any correlation between SARS-Cov2 vaccination on flu incidence and severity in the pediatric population of the Bay Area during 2022 and 2023 flu season.

we intend to analyze if the efforts invested in COVID-19 vaccination are reflected in flu incidence rates and flu severity incidence, investigating whether increased COVID-19 vaccination coverage correlates with a reduction in flu cases and severity in the pediatric population in the Bay Area. If our analysis reveals a lack of correlation, particularly within the pediatric population of Bay Area counties, it emphasizes the necessity for targeted interventions and more outreach efforts should be done to encourage flu vaccinations. 

## Methods
Data Sources, for each data sets are : dataset one (sim_flu_CA.csv) contains weekly data about flu cases and case severity by demographic categories and geographic categories for California counties (except for Los Angeles county) of flu season for California counties for 2022 and -2023 ; dataset two (sim_flu_LACounty.csv) contains population data by similar demographic and geographic categories for Los Angeles county; dataset three (ca_vax_rates_quarter.csv) contains quarterly COVID vaccination rates by county and age category from 2020 to 2023.

Once we downloaded our the data sets, we cleaned the data sets by renaming the sim_flu_LA_county_pop to match the sim_flu_CA names, we then included a  county column in LA dataset in preparation to be added to row of sim_flu_CA and renamed columns to clean the column headers to be all lowercase with no spaces. We continued cleaning for sim_flu_CA and sim_flu_LACounty_pop by selecting matching columns for sim_flu_CA & sim_flu_LACounty_pop. Followed by re-coding, checking for NAs and tidying for matching California data sets. We updated the Date column of sim_flu_LACounty_pop to match the date value of sim_flu_CA. 

We then combined California data sets using cbind function, and chose the demographic strata of interest for final_flu by filtering rows out of selected columns (county, age_cat, dt_diagnosis), population at risk 0-17 years, pediatric population selection of dates prior to flu season and right after flu season 2022- 2023. After reviewing the dates from both data frames, dates of vaccination were selected according to the flu vaccine release, September 2022. Followed by tidying COVID-19 vaccination rates data of interest by pivoting data from ca_vax_rates_quarter to extract the age category. At this point all NULL values were converted to NAs. Defined our at risk population to be 0-17 under one category and removed NAs.

Vaccination data was grouped and filtered by county, year, sex and age_category, and calculated sum of each numeric category and calculated proportion of vaccinated population per county per 100 population.The flu data set was further cleaned to new data frame with the following variables:county, date_diagnosis, age_category, sex, race_ethnicity, new_severe, infected_current, new_recovered, total population. New variables; incidence and prevalence of infection were created. Vaccination period and flu incidence data sets were aligned and dates were converted into day, month, year . At this point incidence for pediatric population and severity of incidence was calculated per county. We created  variables in order to join data sets and visualize our variables. We selected nine counties in the ,please see the data dictionary for a detailed description.

Analytic Methods, we made tables and plots. The first set of our table sare summary descriptive statistics and plot compares the data collected and measures of association obtained between the fourth quarter of 2022 and first quarter of 2023 as well as the vaccination coverage for COVID-19, in the pediatric population . The second set of table and plot compares flu incidence rate per 1000 population with Covid-19 vaccination coverage per 100 population for October - December 2022. The third set of table and plot compares flu severity incidence rate per 1000 population with Covid-19 vaccination coverage per 100 population for October - December 2022. The fourth set of table and plot compares flu incidence rate per 1000 population with Covid-19 vaccination coverage per 100 population for January - March 2023.The fifth set of table and plot compares flu severity incidence rate per 1000 population with Covid-19 vaccination coverage per 100 population for January - March 2023.

## Results

Covid Vaccination Rate Metrics & Summarize
Joining data-sets
After the calculation of flu incidence per 1000 in pediatric population in the staete of    California for the flu season ( 2022 and  2023), and the calculation of vaccination coverage per county, the cleaned sim_ca_la_incidence and ca_vax_rates_grouped were joined.

Vistualizations

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, import statement, tidy_source(), include = FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(tidyverse)
library(lubridate)
#Importing  morbidity data sets for state of California provided in the PHW251 Project Data #github repo

df_ca_vax_rates <- read_csv("https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/scenario_1/ca_vax_rates_quarter.csv")

df_sim_flu_ca <-read_csv("https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/scenario_1/sim_flu_CA.csv")

df_sim_flu_lacounty <-read_csv("https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/scenario_1/sim_flu_LACounty_pop.csv")

```


```{r , include=FALSE,warning=FALSE, message=FALSE}
#Recode column names, values, or formats that are in discordant #(dates, etc)
df1 <- df_sim_flu_ca %>% rename(population=pop, age_category=age_cat) 

df2 <- df_sim_flu_lacounty %>% rename_all(tolower) %>%
  rename(date_diagnosis=dt_dx)

df3 <- df_ca_vax_rates %>% janitor::clean_names() %>% 
  rename(population=estimated_population)

# for data cleaning first use str to have insight of data sets on variable #type
str(df1)
str(df2)
str(df3)

```


```{r , include=FALSE,warning=FALSE, message=FALSE}
# Using dplyr
# Remove duplicate rows (all columns)
library(dplyr)
df1 <- df1 %>% distinct()
df2 <- df2 %>% distinct()
df3 <- df3 %>% distinct()
```


```{r , include=FALSE,warning=FALSE, message=FALSE}
#unique() to identify misspellings in data sets to harmonise spellings
unique(df1$sex)
unique(df2$race_eth)
unique(df2$sex)
unique(df3$suppress_data)
```


```{r , include=FALSE,warning=FALSE, message=FALSE}
#Subset rows or columns, as needed
df1_new <- df1 %>% select(-time_int, -report_date) %>%
  filter (new_infections != 0, new_recovered != 0) 
df2_new <- df2 %>% select (-dt_report) %>% filter(infected_current != 0)

```


```{r , include=FALSE,warning=FALSE, message=FALSE}
#Create new variables needed for analysis (minimum 2)
df1 <- df1_new %>%
  mutate(prevalence=cumulative_infected/population*100,
         incidence=new_infections/population*100)

df2_new <- df2_new %>%
  mutate(prevalence=infected_cumulative/population*100,
         incidence=dx_new/population*100)

```


```{r , include=FALSE,warning=FALSE, message=FALSE}
#re code age_category or sex in df2_new we use unique function
unique(df2_new$age_category)#to identify the diferent age ranges
unique(df2_new$sex)
                               
```

```{r , include=FALSE,warning=FALSE, message=FALSE}
#missing values
# to identify where missing values are
df3_new_missing <-df3%>% filter(!complete.cases(.))
#removing NA for continous variables by droping NA
df3_new_missing <-df3%>%  
  drop_na(population,cumulative_unvaccinated)
sum(is.na(df1))
sum(is.na(df2_new))
sum(is.na(df3_new_missing))#to check if any more missing in df

```

```{r , include=FALSE,warning=FALSE, message=FALSE}
##Re code Male =1 and Female = 2
#df2_new$sex <- ifelse(df2_new$sex == "MALE", 1, 2)
```

```{r , include=FALSE,warning=FALSE, message=FALSE} 
#summary statistics using the describe function from the psych package
library(psych)
describe(df1)
describe(df2_new)
describe(df3_new_missing)
```

```{r , include=FALSE,warning=FALSE, message=FALSE} 
#Summary statistics for individual columns in the dataset
new_infection <- summary(df1$new_infections)
new_infection
new_recover <- summary(df1$new_recovered)
new_recover
count_susceptible <- summary(df1$count_susceptible)
count_susceptible
population <- summary(df1$population)
population
prevalence <- summary(df1$prevalence)
prevalence
incidence <- summary(df1$incidence)
incidence
```

```{r, include=FALSE}
# Rename the columns to ease merging of the two datasets from CA and LA
df2_1 <- df2_new %>%
  rename(
dt_diagnosis = date_diagnosis,
new_infections = dx_new,
new_recovered = recovered_new,
count_susceptible = susceptible,
current_infected = infected_current,
cumulative_infected = infected_cumulative,
cumulative_recovered = recovered_cumulative,
new_severe = severe_new,
cumulative_severe = severe_cumulative
)

#adding a demographic variable (county) to df2_1 (la dataset)

df2_1 <- df2_1 %>% mutate(county = "Los Angeles County")
```


```{r, include=FALSE}
library(lubridate)
# Convert the dt_diagnosis column to the desired format
df2_1$dt_diagnosis <- parse_date_time(df2_1$dt_diagnosis, orders = "dmy")#given that the date is in 03oct2022 format,
df2_new$dt_diagnosis <- format(df2_1$dt_diagnosis, "%Y-%m-%d")
# Verify the updated date format
```


```{r, include=FALSE}
#remove race_ethnicity from both la and ca data
df1 <- df1 %>% select(-race_ethnicity)
df2_1 <- df2_1 %>% select(-race_eth)
```

```{r, include=FALSE}
#q5, data sets from CA and LA are bound together to one dataset
#New combined dataset for ca_la
final_flu <- bind_rows(df1, df2_1)
#splitting up the dt_diagnosis column to get year into one variable
final_flu$dt_diagnosis <- as.Date(final_flu$dt_diagnosis, format = "%d%b%Y")

final_flu <- final_flu %>% 
  mutate(
    yr_dt_dx = format(dt_diagnosis, "%Y"),
    mo_dt_dx = format(dt_diagnosis, "%m"),
    day_dt_dt = format(dt_diagnosis, "%d")
  )
class(final_flu$dt_diagnosis)
final_flu <- final_flu %>% select(-mo_dt_dx,-day_dt_dt)

```


```{r cleaning vax dataset df3 new missing, include=FALSE}
df3_new_missing <- df3_new_missing %>% 
  pivot_wider(names_from = demographic_category,
              values_from = demographic_value, values_fn = list)
#renaming new columns
df3_new_missing <- df3_new_missing %>% 
  rename(sex = Gender,
         age_category = "Age Group",
         race_eth = "Race/Ethnicity")
#remove some columns from vax dataset
df3_new_missing <- df3_new_missing %>%
  select(-county_type, -dt_admin, -suppress_data, -race_eth)
#remove mssing data with na.omit function, verify if missing data still exists with sum(is.na)
na.omit(df3_new_missing)
sum(is.na(df3_new_missing))
#replacing NULL in dataset with NA in columns of interest
df3_new_missing$sex[df3_new_missing$sex == "NULL"] <- NA
df3_new_missing$age_category[df3_new_missing$age_category == "NULL"] <- NA
#to view age categories in df3_new missing("Under 5","5-11","12-17","18-49","50-64","65+")
unique(df3_new_missing$age_category)
#view age group of final_flu dataset ( "0-17"  "18-49" "50-64" "65+" )
unique(final_flu$age_category)

# recode male and rename in df3_missing (vax) dataset to be harmonise with final_flu
df3_new_missing$sex <- ifelse(df3_new_missing$sex == "male", "MALE", "FEMALE")
#recode df3_missing(vax dataset) age group to be harmonised with flu dataset agegroup
df3_new_missing$age_category <- ifelse(df3_new_missing$age_category %in% c("under 5", "5-11", "12-17"), "0-17", 
                                       ifelse(df3_new_missing$age_category %in% c("18-49", "50-64", "65+"), df3_new_missing$age_category, NA))
sum(is.na(final_flu))
sum(is.na(df3_new_missing))
#rename county_name in df3_new dataset
df3_new_missing <- df3_new_missing %>% rename(county = county_name)
df3_new_missing$age_category <- unlist(df3_new_missing$age_category) # to convert age_category
#from list to character
#splitting up the dt_diagnosis column to get year into one variable in df3 dataset
df3_new_missing$quarter <- as.Date(df3_new_missing$quarter, format = "%d%b%Y")
```


```{r, include=FALSE}
df3_new_missing <- df3_new_missing %>% 
  mutate(
    yr_dt_vax = format(quarter, "%Y"),
    mo_dt_vax = format(quarter, "%m"),
    day_dt_vax = format(quarter, "%d")
  )
class(df3_new_missing$quarter)
df3_new_missing <-  df3_new_missing %>% select(-mo_dt_vax,-day_dt_vax) %>%
  mutate(vacc_rate = (cumulative_fully_vaccinated/population)*100)
```

Table 1: Summary statistics of California flu showing demographic variables and severity for 2022 and 2023

From Table 1, the proportion of females and males in the different age groups was almost the same. The proportion of men exceeded that of women in the age group 18-49 by 1% and the proportion of females within the age group 65+ exceeded that of men by 1%. The highest number of new infections were experienced within the age group 18-49. The recovery rate was highest within the age group 18-49. The severity of the virus was highest within the age groups 50-64 and 65+. 
Vaccination rate showed to have increased in 2023 as compared to 2022 within all the different age groups.

```{r, echo=FALSE, message=FALSE}
library(gtsummary)
#filtering rows out of selected columns (county, age_category, dt_diagnosis), population at risk #0-17 years
unique(final_flu$county)
subset_final_flu <- final_flu %>% select(sex, age_category, incidence, dt_diagnosis, county) %>% filter(age_category== "0-17" & sex== "MALE" )
#subset_final_flu %>% select(county %in% c("Alameda County","Contra Costa County","San Francisco County","Los Angeles","Sacramento County"))
#summary statistics of joint California flu dataset (final_flu)
 tbl_summary_1 <- final_flu %>%
  select(sex, new_infections, prevalence, new_recovered, new_severe, current_infected, yr_dt_dx, new_recovered,count_susceptible, cumulative_severe, cumulative_recovered, new_unrecovered, age_category) %>% tbl_summary(by=age_category, statistic = list(
    all_continuous() ~ "{mean} ({sd})",
    all_categorical() ~ "{n} / {N} ({p}%)"),
  )
 tbl_summary_1
```
Table 2: Summary descriptive statistics of COVID vaccination in California between 2020-2023

```{r echo=FALSE, message=FALSE, warning=FALSE}
# summary statistics for vaccination dataset (df3_new_missing)
library(formattable)
#to join all 3 data sets
subset_df3_new_missing <- df3_new_missing %>% select(age_category, sex, vacc_rate,county)  %>% 
filter(age_category == "0-17") 
join_flu_vax <- left_join(subset_final_flu, subset_df3_new_missing,)

df3_new_missing %>% summarise(mean_vacc=mean(vacc_rate),
                              mean_cum_ful_vac=mean(cumulative_fully_vaccinated),
                              sd_= sd(cumulative_fully_vaccinated),
                              range_year=range(yr_dt_vax),
                              min_cum_todate_vax=min(cumulative_up_to_date_vax),
                              max_cum_uptodate_vax=max(cumulative_up_to_date_vax))
describe(df3_new_missing)
  
```

Chart: Flu incidence rate in pediatric population of the Bay Area during the Flu season of 2022 and 2023
```{r echo=FALSE, message=FALSE}
library(dplyr)
ggplot(final_flu, aes(x= county, y= incidence)) +
    geom_bar(stat = "identity", color = "blue", fill = "blue") +
  labs(x = "County", y = "incidence" ,
   title = "Flu incidence in paediatrics age 0-17 years for California counties , 2022-2023") +
  theme(axis.text.x = element_text(size = 6, angle = 45)) #+theme_bw()

?geom_bar
```
```{r echo=FALSE, message=FALSE, warning=FALSE}

```

```{r echo=FALSE, message=FALSE}
#Avoid overcrowding of the virsualization of county by incidence
# Create two dataframes for counties starting with letters A-M and N-Z
counties_AtoM <- final_flu %>% filter(substr(county, 1, 1) %in% LETTERS[1:13])
counties_NtoZ <- final_flu %>% filter(substr(county, 1, 1) %in% LETTERS[14:26])

# Visualization for counties A-M
ggplot(counties_AtoM, aes(x = county, y = incidence)) +
  geom_bar(stat = "identity", color = "orange", fill = "orange") +
  labs(x = "County", y = "incidence",
       title = "Flu incidence in paediatrics age 0-17 years for California counties (A-M), 2022-2023") +
  theme(axis.text.x = element_text(size = 6, angle = 45))
```

From the figure above, flu incidence was lowest in Alameda County which reported about 5000 new cases and was highest in Colusa County, Del Norte County, Inyo County, Lassen County, and Monterey County which reported about 24000 new cases per total population.

```{r echo=FALSE, message=FALSE}
# Visualization for counties N-Z
ggplot(counties_NtoZ, aes(x = county, y = incidence)) +
  geom_bar(stat = "identity", color = "darkgreen", fill = "darkgreen") +
  labs(x = "County", y = "incidence",
       title = "Flu incidence in paediatrics age 0-17 years for California counties (N-Z), 2022-2023") +
  theme(axis.text.x = element_text(size = 6, angle = 45))
```
```{r echo=FALSE, message=FALSE}
# Visualization for counties A-M with different fill colors for each county
ggplot(counties_AtoM, aes(x = county, y = incidence, fill = county)) +
  geom_bar(stat = "identity", color = "blue") +
  labs(x = "County", y = "incidence",
       title = "Flu incidence in paediatrics age 0-17 years for California counties (A-M), 2022-2023") +
  theme(axis.text.x = element_text(size = 6, angle = 45)) +
  scale_fill_manual(values = rainbow(nrow(counties_AtoM)))

# Visualization for counties N-Z with different fill colors for each county
ggplot(counties_NtoZ, aes(x = county, y = incidence, fill = county)) +
  geom_bar(stat = "identity", color = "blue") +
  labs(x = "County", y = "incidence",
       title = "Flu incidence in paediatrics age 0-17 years for California counties (N-Z), 2022-2023") +
  theme(axis.text.x = element_text(size = 6, angle = 45)) +
  scale_fill_manual(values = rainbow(nrow(counties_NtoZ)))
```






Discussion


The new infection rate of the virus was highest in the age group 18-49 years. This could be due to the fact that, this age group is the most active age group. The mobility of the persons might have exposed more to infection as compared to their less active younger and older age group counterparts. The severity of the virus was shown to be more in the older age groups. This could be due to the fact that, age group usually comes with increase cormodities from different non communicable diseases which have been shown to be positively associated to severity of the virus.
In the obtained data from the Bay Area Counties in pediatric population during the observed period (Oct-2022 to Mar-2023), there was an increase of 3.4 times in the flu cases despite the same proportion of coverage of COVID-19 vaccination in the Bay Area with no significant differences in morbidity.

Our analysis revealed no relation between COVID-19 vaccination and flu cases in the Bay Area counties in pediatric population. Despite the exponential increase of cases between the Q4 of 2022 to the Q1 of 2023, the incidence of severe cases was similar.
Our findings demonstrates the importance of implementing intensified outreach efforts for flu vaccination in the pediatric population of the Bay Area, with a strategic emphasis on targeted interventions to enhance coverage and mitigate the impact of influenza in this demographic.

Conclusion


A coordinated effort in required from multidisciplinary sectors to ensure the virus incidence is greatly reduced and subsequently eradicated.


