---
title: "Problem set 1"
author: "Ukah Chrisantus Eweh, Ngu Claudia Ngeha"
date: "2023-10-02"
output: html_document
---


##Description of dataset
#What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)
The data is simulated data that represents Flu outbreak and vaccine uptake for the entire state of California during fall 2023.This scenario contains demographic (age category, gender, race/ethnicity), geographic (county), morbidity, disease severity, and county and regional vaccination data from the simulated flu season.

#How does the dataset relate to the group problem statement and question?
The datasets contains the different variables necessary for calculation of vaccination rate for flu and to carryout the comparative analysis with that of Covid vaccine.

#Import statements for all datasets

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
ca_vax_rate <- read_csv("https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/scenario_1/ca_vax_rates_quarter.csv")
ca_vax_rate

#Utilize function arguments to control relevant components (i.e. change column types, column names, missing values, etc.)

df <- ca_vax_rate %>%
rename_all(tolower) %>% # Convert column names to lowercase
rename_all(~gsub(" ", "", .)) %>% # Remove spaces from column names
rename(population = estimatedpopulation)
df

```

```{r}
sim_flu_ca <- read_csv("https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/scenario_1/sim_flu_CA.csv")
```

```{r}
sim_flu_la <- read_csv("https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/scenario_1/sim_flu_LACounty.csv")
```

#Document the import process
The process was as foolows;
 -PHW251 repository was opened from github
 -The scenario(scenario 1 dataset) was opened
 -Each of the dataset to be imported in the scenario was opened to get the url
 link
 -We clicked on "Raw" and the url of the dataset appeared on the browser which was then copied
 -Rstudio was opened and the library readr was loaded
 -The function read_csv was used to load the data into 
 as follows;
 library(readr)
 dataset <- read_csv (datasetpath)
 