---
title: "Project Milestone 2"
author: "Ukah Chrisantus Eweh, Ngu Claudia Ngeha"
date: "2023-10-02"
output: html_document
---


##Description of dataset
#What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)
The data is simulated data that represents Flu outbreak and vaccine uptake for the entire state of California during fall 2023.This scenario contains demographic (age category, gender, race/ethnicity), geographic (county), morbidity, disease severity, and county and regional vaccination data from the simulated flu season.

#How does the dataset relate to the group problem statement and question?
The datasets contains the different variables necessary for calculation of vaccination rate for flu and to carryout the comparative analysis with that of Covid vaccine.

#Import statements for all datasets

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
ca_vax_rate <- read_csv("https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/scenario_1/ca_vax_rates_quarter.csv")
ca_vax_rate

#Utilize function arguments to control relevant components (i.e. change column types, column names, missing values, etc.)

df <- ca_vax_rate %>%
rename_all(tolower) %>% # Convert column names to lowercase
rename_all(~gsub(" ", "", .)) %>% # Remove spaces from column names
rename(population = estimatedpopulation)
df

```

```{r}
sim_flu_ca <- read_csv("https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/scenario_1/sim_flu_CA.csv")
sim_flu_ca
```

```{r}
sim_flu_la <- read_csv("https://raw.githubusercontent.com/PHW290/phw251_projectdata/main/scenario_1/sim_flu_LACounty.csv")
sim_flu_la
```

#Document the import process
The process was as foolows;
 -PHW251 repository was opened from github
 -The scenario(scenario 1 dataset) was opened
 -Each of the dataset to be imported in the scenario was opened to get the url
 link
 -We clicked on "Raw" and the url of the dataset appeared on the browser which was then copied
 -Rstudio was opened and the library readr was loaded
 -The function read_csv was used to load the data into 
 as follows;
 library(readr)
 dataset <- read_csv (datasetpath)
 
 
 #CLEAN UP COLUMNS NAMES 
```{r}
# Get the current column names for ca_vax_rate dataset
col_names <- colnames(ca_vax_rate)

# Clean up column names using snake case
clean_col_names <- gsub("[^[:alnum:]]", "_", tolower(col_names))

# Assign the cleaned column names back to the dataset
colnames(ca_vax_rate) <- clean_col_names
print(ca_vax_rate)
# Now the column names are in snake case
```
```{r}
# Get the current column names for sim_flu_ca dataset
col_names <- colnames(sim_flu_ca)

# Clean up column names using snake case
clean_col_names <- gsub("[^[:alnum:]]", "_", tolower(col_names))

# Assign the cleaned column names back to the dataset
colnames(sim_flu_ca) <- clean_col_names
print(sim_flu_ca)
# Now the column names are in snake case
```
```{r}
# Get the current column names for sim_flu_la dataset
col_names <- colnames(sim_flu_la)

# Clean up column names using snake case
clean_col_names <- gsub("[^[:alnum:]]", "_", tolower(col_names))

# Assign the cleaned column names back to the dataset
colnames(sim_flu_la) <- clean_col_names
print(sim_flu_la)
# Now the column names are in snake case
```
 
 
##IDENTIFY DATA TYPES FOR 5+ DATA ELEMENTS/COLUMNS/VARIABLES
 
```{r}
str(ca_vax_rate) #Data types for the ca_vax_rate dataset
```
 
```{r}
str(sim_flu_la) #Data types for the sim_flu_la dataset
```
```{r}
str(sim_flu_la) #data types for the sim_flu_la dataset
```
 
 UTILIZE FUNCTIONS OR RESOURCES IN RSTUDIO TO DETERMINE THE TYPES OF EACH OF EACH DATA ELEMENT
 
```{r}
sapply(ca_vax_rate, class) #sapply() is used along with class() to determine the types of all variables in a dataset
```
```{r}
sapply(sim_flu_la, class)
```
```{r}
sapply(sim_flu_la, class)
```
 
##will you need to convert any columns to numeric or another type?
 
Yes we might convert sex to factor (numeric;1,2) using as.factor. This will the variable "sex" to actually in levels/category and here will be able to clearly how the out varies in one sex category as compared to the others.
asked to run otherwise we could still leave sex as character variable (female,male).


#PROVIDE A BASIC DESCRIPTION OF THE 5+ DATA ELEMENT

```{r}
# Numeric data elements for the ca_vax_rate dataset
numeric_cols <- sapply(ca_vax_rate, is.numeric)  # Identify numeric columns
numeric_summary <- data.frame(
  Mean = sapply(ca_vax_rate[, numeric_cols], mean),
  Median = sapply(ca_vax_rate[, numeric_cols], median),
  Range = sapply(ca_vax_rate[, numeric_cols], function(x) max(x) - min(x))
)

# Character data elements
character_cols <- sapply(ca_vax_rate, is.character)  # Identify character columns
character_summary <- data.frame(
  Unique_Values = sapply(ca_vax_rate[, character_cols], function(x) length(unique(x)))
)

# Print the summaries
print(numeric_summary)
print(character_summary)


#In this code, `sapply()` is used to apply the respective functions (`mean()`, `median()`, `function(x) max(x) - min(x)`, and `function(x) length(unique(x))`) to each numeric and character column in the dataset. The results are stored in separate data frames (`numeric_summary` and `character_summary`), and then printed.
```

#Please kindly insert the solution for description of 5+ data elements here


#Describe cleaning that each data source may need
-In the dataset ca_vax_rate, the columns; demographic_category and demographic levels needs cleaning and
re-categorization.
• When using the datasets ( all three data sets), they have to be cleaned by removing “NA” (missing
variables) using logic functions.
• Rate of flu vaccination column has to be calculated and inserted in the data set for flu vaccination
using the mutate function.
• The two data sets ca_vax_rate(for covid vaccinnation rate) and the dataset, sim_flu_ca needs to be
joint for correlation analysis to be done.
• Sex variable needs to be converted from character variable (male and female) to as.factor (1,2) for easy
manipulation in data set
